A look into Kamala Harris' polls in the upcoming 2024 US Presidential Election 

### 2024-10-25 paper/paper.qmd
Uploaded progress as of 25/10/2024
Progress just before agreeing to collaborate on project. Parts of the introduction, abstract, and data and model sections are completed, along with a rudimentary model based upon pollster data predicting a Harris win.

---
title: "USA pollster"
subtitle: "CHANGE SUBTITLE"
author: 
  - Chris Yong Hong Sen
thanks: "Code and data are available at: [https://github.com/Monoji77/USA-pollster](https://github.com/Monoji77/USA-pollster)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
library(knitr)
```

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

@@

Weaknesses and next steps should also be included.

@@

\appendix

# Appendix {-}

# Patriot Polling: Wisconsin Presidential Analysis Data Collection

## Methodology
From 12 to 14 October 2024, Patriot Polling has conducted phone surveys in the state of Wisconsin of the United States of America, which borders Minnesota, Iowa, Michegan, and Illinois. Their target population are voters in Wisconsin. For sample of landlines which are characterised by households, an equal number of phone numbers are randomly selected across every landline block in Wisconsin using Random Digit Dialing (RDD). As for the sample of mobile numbers, which are characterised by an individual with access to a digital mobile device, the sample was purchased from a consumer contact number database. For this poll, a total of 803 respondents has completed the survey either from contact through the landline or their personal contact number. [@patriotPollingMethodology]

@patriotPollingMethodology conducted polling in such a way where a randomly contacted respondent would hear pre-recorded voice messages and users are able to interact with the automated phone system. This system is called Interactive voice response (IVR) which can handle outbound calls more systematically since the same questions would be asked sequentially in the survey. There were only 2 questions that were asked. Both questions begin with 'How will you vote for president?'. This is followed by the candidates name as it appears in the voting ballot.

## Strengths
This automated method of phone surveys is a lot more convenient and efficient than in person type of surveys since it removes the need for a surveyor to visit respondents face to face and also removes the need for a surveyor to contact multiple respondents over the phone. This decreases labour effort and cost.

Furthermore, with real-time feedback from phone surveys, the list of questions can be easily and quickly adjusted to ensure the ethical integrity of the survey being conducted. According to @patriotPollingMethodology, this survey took place in 2 daysm which could have taken a lot longer if other in-person type of surveying methods were used in this research.

Particularly, the polling company used a repertoire of RDD sampling and stratified sampling on landlines to ensure that every landline block (strata) has equal numbers of respondents to be included in the final sample. This helps to ensure diversity from the sample by including different geographical landline blocks, reducing sampling bias. The purchasing of mobile phones from phone number databases abates sampling bias in areas where a particular demographic - younger people who are fully dependant on mobile phones and not landlines - would otherwise be underrepresented if only RDD sampling was conducted on landlines.

## Limitations and weakness
Firstly, there was a lack of information regarding how RDD was performed for the landline sample, how they handled non-working landlines, as well as the total sample size attributed from landline and mobile phone surveys.

The handling of no-response entities were also not explicitly mentioned by Patriot Polling. This could translate to respondents who were chosen but did not pick up their mobile phones/landlines as it was a busy period in the day or due to potential heightened awareness of phone based scams. This means that the final sample obtained might not be representative of the Wisconsin voter population.

Also, the initial phrasing to the second question can very easily confuse respondents. Although the initial phrasing is followed by the senate candidate's name, the initial phrasing 'How will you vote for president' still uses the wrong word 'president' instead of 'senate'. This does not reflect a high quality survey, which could lead to decreased trust from the respondent to Patriot Polling, in turn potentially resulting in a premature end to the surveys, voiding the initial response.

Lastly, the purchasing of phone numbers from a consumer database would introduce of selection bias. Individuals from middle to higher socio-economic status would have greater purchasing power than their lower income counterparts, which means individuals from lower socio-economic status may not have phones. In addition, elderlies within Wisconsin may not necessarily have mobile phones due to the lack of skills to use such technologies. Also, individuals would've opted for do-not-call registries, opted out of marketing databases or simply have prepaid phones which do not link to their personal information. This could mean an under representation of individuals from the said groups.

```{r senate they voted for}
#| label: tbl-senate
#| include: false
#| echo: false
#| layout-ncol: 2
#| fig-cap: "Presidental Candidates Votes"
#| fig-subcap: ["Percentage of votes for senate candidates in Wisconsin", "Percentage of votes for presidential candidates in Wisconsin"]

presidential_voter_table <- tibble(
  'Presidential Candidate'= c('Donald Trump (R)', 'Kamala Harris (D)'),
  'Registered Voters'= c('50%', '49%'),
)
senate_voter_table <- tibble(
  'Senate Candidate'= c('Tammy Baldwin (R)', 'Eric Hovde (D)'),
  'Registered Voters'= c('50%', '49%'),
)

kable(presidential_voter_table) 
kable(senate_voter_table)
```

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

---
title: "A look into Kamala Harris' polls in the upcoming 2024 US Presidential Election "
subtitle: ""
author: 
  - Veyasan Ragulan
thanks: "Code and data are available at: https://github.com/Veyasan1/2024USElectionModel"
date: today
date-format: long
abstract: "The 2024 US election has been seen as a close race between the Democratic nominess Kamala Harris, and the Republican Nominee Donald Trump, with each candidate holding down key states and matching each other in battleground states such as Pennsylvania and Georgia. This paper looks at the polling as of late October 2024 to see how Kamala Harris fares as the election draws near."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)

bayesian_model <-
  readRDS(file = here::here("models/electionmodel.rds"))

spline_model <- 
  readRDS(file = here::here("models/splinemodel.rds"))
```

@@

In this paper we look at polling data from 538 on the 2024 US Presidential election, specifically on Demoncratic Candidate Kamala Harris. We will construct a model that will be used to make a prediction on whether Harris wins the US Presidency

Section 2 will outline the source of this data. Section 3 covers the model and it's parameters. Section 4 is where discussion will be made about the models predictions and how realistically they line up with current affairs. Finally, section 4 discusses any weakness and limitations that can be considered for another report.

This report was written with the assistance of R @citeR. Additionally, @tidyverse, and @rstanarm were used to clean the data, write models, and analyse results. The structure of this report heeds Rohan Alexander's example @tellingstories.

@@

The data used in this model comes from fivethirtyeight.com, a polling site known for their in-depth analysis and data on US elections @fivethirtyeight. Here, we have used the 2024 general election data, originally accessed on 10/21/2024. The data contains information from several polls taken across the United States, including the operator, sponsors, methodology, a numeric grade to quanitfy the quality of the poll, and when and where the poll was taken. From this dataset, polls predicting a Harris win were considered, to construct the predictive model.

@@

The model contains 4 parameters. They are pollster, state, sponsors, and methodology.

Pollster is the operator of the observed poll. Pollster provides a relatively easy way to identify trends in certain polls, depending on who is conducting them.

State tells us where the poll was conducted. Across the United States, ther are states that traditionally vote Democrat (New York, California), some that traditionally vote Republican (Texas, Florida), and 'swing states' (Michigan, Pennsylvania). Swing states in particular are the most interesting states to watch as votes from these states tend to win a candiate the US presidency.Knowing where the polls have been taken also lets us understand if their choice was one bound by the state's culture.

Sponsors are organizations that fund or otherwise assist the pollster in operating the poll. Sponsors carry the risk of bias with them, potentially influencing the pollster, and thus the polls to skew in a direction the sponsor favors preferably. Knowing who these sponsors are potemntially lets us make informed predictions on the outcome of polls.

Methodology is the way the poll was conducted. Sometimes the composition of the poll influences peoples votes, so this is an interesting parameter to track. This also lets us see if certain methods yield better or worse predictions.

@@

```{r}
#| echo: false
#| warning: false
#| message: false

# Posterior predictive checks
pp_check(bayesian_model)

# Summarize the model
summary(bayesian_model)

# Plot random effects
plot(bayesian_model, pars = "(Intercept)", prob = 0.95)

```

```{r}
#| label: fig-1
#| fig-cap: Street tree count of 10 most popoulus tree species in Toronto up to September 2024
#| echo: false
#| warning: false
#| message: false

# Summarize the model
summary(spline_model)

# Posterior predictive checks
pp_check(spline_model)
```

@@

## Trends from pollsters

## Trends from states

## Trends from sponsors

## Trends from methodology

@@

@@

# Appendix

## Pollster Analysis

## Poll Blueprint
